Comparing the Jacobi Method and LLL lattice reduction algorithms for cryptographic applications

### 2014-12-19 report/paper.tex
19/12 update

@@

\section{Introduction}
\subsection{Results and Contributions}

\my{We show the following main results:
\begin{itemize}
\item An analysis of running times depending on factors
\item paper: \cite{originalJacobiMethodLatticeBasisReduction}
\end{itemize}
}
\subsection{Related Work}

@@

@@

\begin{itemize}
\item the diagonal has 1 for all it's entries except one position that is set to a prime number $P$. Hence, the $\det(B)$ is prime.
\item All column entries of the matrix under the position that is set to $P$ are smaller than $P$ in absolute value.
\end{itemize}

Without loss of generality, we hence restrict tests to random lattices of volume $P$ whose basis in HNF form is as follows:
$$\begin{array}{ccccc}
P & \vec{a_2} &  & \\
\vec{a_2} & 1&   & \\ 
\dots & &  & \\ 
\vec{a_{m}} & & &1 
\end{array}$$
where $a_i \in \mathbb{Z}/ P\mathbb{Z}$.

@@

The orthogonality defect is a quality indicator that we will use to compare the reduction of \emph{the set of basis vectors}. The orthogonality defect can be interpreted geometrically \my{you recite the definition, no geometrical interpretation is given??} as the product of the basis vector lengths divided by the volume of the lattice. As the basis vectors are getting increasingly orthogonal during a reduction, the nominator will become smaller. The lower bound of the orthogonality defect is $\text{OrthDefect} \ge 1$. In the special case where the basis would be completely orthogonal, the orthogonality defect would be $\text{OrthDefect} = 1$.

@@

From the definition of a Lagrange-reduced basis, we can define a Jacobi-reduced basis \my{replace 'such as' by 'for which'}:
\begin{align}
\langle \vec{b_i} \vec{b_j} \rangle &\leq \frac{1}{2} \| \vec{b_i}\|^2 \text{ (for all $1 \leq i < j \leq n$)}.\\
\| \vec{b_i}\| &\leq \|\vec{b_j} \| \text{ (for all $1 \leq i < j \leq n$)}
\end{align}

@@

\subsubsection{$\omega$-Lagrange-reduced conditions}
An \my{A parameter} $\omega$ is introduced to \my{...? }
There are two conditions for a basis to be $\omega$-Lagrange-reduced.

\[
\begin{cases}
\lfloor g_{ij} / g_{ss} \rceil \leq 1, \\
\omega^2 g_{ll} \leq g_{ii} + g_{jj} - 2|g_{ij}|

\end{cases}
\]  where $1/\sqrt{3} \leq \omega < 1$ \my{and where $g_{ij}$}.. .

@@

\input{algorithms/iterativeLangrange}

@@

We implemented the \my{variant on the} Jacobi reduction in C++ \my{by use of} with the \emph{newNTL} library. The FPLLL implementation \my{of LLL} was used \my{to compare the results after reduction.}

@@

Our experimentation focused on the quality of the reduction. The two quality \my{metics} we used for comparing the reductions are:

@@

\subsection{Input bases}
Numerous papers were written about the Jacobi method for lattice reduction by S. Qiao \cite{originalJacobiMethodLatticeBasisReduction}\cite{complexityAnalysisOfJacobiMethod}\cite{enhancedJacobi} but none of the papers are clarifying what bitsize was used for the randomly generated entries of the vectors. The main focus of the authors was the use of lattice reduction for signal processing, more specifically multiinput and multioutput (MIMO), the use of multiple antennas at both the transmitter and receiver to improve communication reliance. MIMO uses smaller dimensions lattices. \my{In} \cite{enhancedJacobi} and \cite{tian2014hybrid} \my{a reduction is performed on} lattices of dimensions 20 to 100. These dimensions are inappropriate for cryptographic applications.

In our experiments, we have used a bitsize of around 10 bits entries when \my{actually always} generating the almost orthogonal bases.

@@

We initially ran our Jacobi reduction variant on almost orthogonal bases to attempt to reproduce Qiao's results.
The orthogonality defects are similar to \my{what} the authors found in \cite{tian2014hybrid}. \my{Our results are shown in Fig.~\ref{fig1} and Fig.~\ref{fig2}. We vary $\omega$..... We observe that...}

\begin{figure}[H]
\begin{minipage}{.5\textwidth}

  \centering
    \includegraphics[width=0.9\textwidth]{random-matrix-omega-06defect.png}
      \caption{LLL with delta $\delta = 0.99$ and Jacobi with $\omega = 0.6$}
  \label{fig1}
\end{minipage}
\begin{minipage}{.5\textwidth}

  \centering
    \includegraphics[width=0.9\textwidth]{random-matrix-omega-09defect.png}

  \caption{LLL with delta $\delta = 0.99$ and Jacobi with $\omega = 0.99$}
  \label{fig2}
\end{minipage}

\end{figure}

\subsection{Random Lattices}

TODO:

@@

We notice that the Jacobi algorithm does reduce the bases further than LLL.

We show two examples of bases that are LLL-reduced but not $\omega > 0.9$ Jacobi reduced.

@@

\section{Conclusions}
\my{Short paragraph}
\section{Credits \my{ or Acknowledgment}}

>>>>

\usepackage{tikz}
\usetikzlibrary{shapes,decorations.pathreplacing,calc,shapes.callouts,decorations.pathmorphing}

\usepackage[normalem]{ulem}
\newcommand{\tikzmark}[1]{\tikz[overlay,remember picture] \node (#1) {};}
\definecolor{mygreen}{rgb}{.222,.62,.49}

@@

\section{Introduction}
In 2012, Sanzheng Qiao published a paper\cite{originalJacobiMethodLatticeBasisReduction} proposing the use of the Jacobi method for lattice reduction applications. In following papers, Qiao proposed several enhancements to the generic Jacobi method. We implemented the Jacobi method reduction algorithm and compared the quality of the outputs with the LLL algorithm.

@@

\begin{center}
\input{figlat/latticegrid}\hspace{2cm}
\input{figlat/latticeZ}
\end{center}

@@

\begin{itemize}
\item the diagonal has 1 for all its entries except one position that is set to a prime number $P$. Hence, the $\det(B)$ is prime.
\item All row entries of the matrix right to the position that is set to $P$ are smaller than $P$ in absolute value.
\end{itemize}

Without loss of generality, we hence restrict tests to random lattices of volume $P$ whose basis in HNF form is as follows:
$$\begin{array}{ccccc}
P & \vec{a_2} & \dots & \vec{a_{m}}\\
& 1&   & \\ 
& & \ddots & \\ 
 & & &1 
\end{array}$$
where $a_i \in \mathbb{Z}/ P\mathbb{Z}$.

@@

The orthogonality defect is a quality indicator that we will use to compare the reduction of \emph{the set of basis vectors}. As the basis vectors are getting increasingly orthogonal during a reduction, the nominator will become smaller. The lower bound of the orthogonality defect is $\text{OrthDefect} \ge 1$. In the special case where the basis would be completely orthogonal, the orthogonality defect would be $\text{OrthDefect} = 1$.

@@

From the definition of a Lagrange-reduced basis, we can define a Jacobi-reduced basis for which:
\begin{align}
\langle \vec{b_i} \vec{b_j} \rangle &\leq \frac{1}{2} \| \vec{b_i}\|^2 \text{ (for all $1 \leq i < j \leq n$)}.\\
\| \vec{b_i}\| &\leq \|\vec{b_j} \| \text{ (for all $1 \leq i < j \leq n$)}
\end{align}

@@

\subsubsection{$\omega$-Lagrange-reduced conditions}
There are two conditions for a basis to be $\omega$-Lagrange-reduced expressed with the Gram matrix $G$ of the basis $B$. The Gram matrix $G$ of a basis $B$ is computed $G = B^T B$. By construction, a Gram matrix is always symmetric, for every pair of indices $(i,j)$, we have $g_{ij} = g_{ji}$.

\[
\begin{cases}
\lfloor g_{sl} / g_{ss} \rceil \leq 1, \\
\omega^2 g_{ll} \leq g_{ss} + g_{ll} - 2|g_{sl}|

\end{cases}
\]  where $1/\sqrt{3} \leq \omega < 1$. The $s$ and $l$ indices are referring to the index of the short and long vectors in the basis Gram matrix of the basis.

@@

\input{algorithms/lagrangeIT-full}

@@

We implemented the Fast-Jacobi\cite{fastJacobi} Jacobi reduction in C++ on top of the \emph{newNTL} library. The FPLLL implementation of LLL was used to compare the results after reduction.

@@

Our experimentation focused on the quality of the reduction. The two quality metics we used for comparing the reductions are:

@@

\subsection{Input bases}
Numerous papers were written about the Jacobi method for lattice reduction by S. Qiao \cite{originalJacobiMethodLatticeBasisReduction}\cite{complexityAnalysisOfJacobiMethod}\cite{enhancedJacobi} but none of the papers are clarifying what bitsize was used for the randomly generated entries of the vectors. The main focus of the authors was the use of lattice reduction for signal processing, more specifically multiinput and multioutput (MIMO), the use of multiple antennas at both the transmitter and receiver to improve communication reliance. MIMO uses smaller dimensions lattices. In \cite{enhancedJacobi} and \cite{tian2014hybrid}, a reduction is performed on lattices of dimensions 20 to 100. These dimensions are inappropriate for cryptographic applications.

In our experiments, we have used a bitsize of around 10 bits entries when actually always generating the almost orthogonal bases.

@@

We initially ran our Jacobi reduction variant on almost orthogonal bases to attempt to reproduce Qiao's results. The orthogonality defects are similar to what the authors found in \cite{tian2014hybrid} and \cite{fastJacobi}. Our results are shown in Fig.~\ref{fig1} and Fig.~\ref{fig2}. By increasing the parameter $\omega$, we are making the $\omega$-Lagrange-reduced conditions more restrictive. More specifically, for every pair of vectors of the basis, the following condition needs to be satisfied
\[
 \omega \| \vec{a}_l \| \leq \|\vec{a}_l \pm \vec{a}_s \|
\] where $\vec{a}_l$ is the longest vector and $\vec{a}_s$ is the shortest. Hence, the reduction renders the basis more orthogonal as $\omega$ grows.

\begin{figure}[H]
\begin{minipage}{.45\textwidth}
  \centering
    \includegraphics[width=0.9\textwidth]{results-graphs/random-matrix-omega-06defect.png}
      \caption{LLL with delta $\delta = 0.99$ and Jacobi with $\omega = 0.6$}
  \label{fig1}
\end{minipage}
\hspace{.05\textwidth}
\begin{minipage}{.45\textwidth}

  \centering
    \includegraphics[width=0.9\textwidth]{results-graphs/random-matrix-omega-09defect.png}

  \caption{LLL with delta $\delta = 0.99$ and Jacobi with $\omega = 0.99$}
  \label{fig2}
\end{minipage}
\end{figure}

The increase in the parameter $\omega$ comes at a significant performance cost as illustrated in Fig~\ref{fig3}.

\begin{figure}[H]
\centering
\includegraphics[width=0.5\textwidth]{results-graphs/runningtime.png}
\label{fig3}
\caption{Running time increasing substantially with $\omega$ parameter}
\end{figure}

\subsection{On Random Lattices}

Our experiments on random lattices on researched dimensions ($> 50$) have been aborted due to excessive running times.

@@

As we found that Jacobi could reduce a random lattice further than LLL when $\omega$ is close to 1, we tried to run the Jacobi reduction algorithm after LLL reduction of a random lattice to see if it was able to reduce it further. Experimental results (Fig.~\ref{fig4}-\ref{fig5}) are showing that the Jacobi method does indeed reduces the base further.

\begin{figure}[H]
\begin{minipage}{.45\textwidth}
  \centering
    \includegraphics[width=0.9\textwidth]{results-graphs/LLLThenJacobi/defect.png}
      \caption{Jacobi after LLL}
  \label{fig4}
\end{minipage}
\hspace{.05\textwidth}
\begin{minipage}{.45\textwidth}

  \centering
    \includegraphics[width=0.9\textwidth]{results-graphs/LLLThenJacobi/defect-logscale-difference.png}

  \caption{Logarithmic-scale difference in orthogonality defects}
  \label{fig5}
\end{minipage}
\end{figure}

We notice that the Jacobi algorithm does reduce the bases further than LLL. Looking at the conditions to be $\delta$-LLL reduced and $\omega-$Jacobi reduced, we can find two bases that are LLL reduced but not Jacobi reduced.

@@

\section{Further research}
As follow-up of this research, we believe it would be useful to:
\begin{itemize}
\item determine the algorithmic complexity of the Jacobi method
\item benchmark the Jacobi method against BKZ after LLL
\end{itemize}

\section{Conclusions}

Although the Jacobi method might have practical applications in signal processing, it is not appropriate for cryptographic applications on random lattices or higher dimension bases.

\section{Acknowledgments}
I would like to particularly thank Anja Becker and Nicolas Gama from LACAL who have been providing significant feedback and advice during my research.
